---
title: "Choropleth Map Project"
subtitle: SDS 264 Spring 2026 - Mini Project 3
title-block-banner: true
---

```{r}
#| message: false
#| warning: false
#| include: false

library(tidyverse)
library(mdsr)
library(maps)
library(lubridate)
library(statebins)
library(sf)
library(poliscidata) 
library(htmltools)
library(glue)
library(leaflet)
states <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson")
```

```{r}
#| message: false
#| warning: false
#| include: false

# Make sure all keys have the same format before joining:
# all lower case, no internal or external spaces

# Cleaning.....
library(maps)
states_polygon <- as_tibble(map_data("state")) |>
  select(region, group, order, lat, long)

library(sf)
states_sf <- 
  read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") |>
  select(name, geometry)

states_polygon
states_sf

my_polisci_data <- as_tibble(poliscidata::states) |>
  select(state, prolife, gay_support3)

my_polisci_data <- my_polisci_data |>
  mutate(
    state = str_to_lower(state),
    state = str_squish(state)
  )

states_sf <- states_sf |>
  mutate(
    name = str_to_lower(name))

states_polygon <- states_polygon |> 
  filter(!region %in% c("alaska", "hawaii"))
states_sf <- states_sf |> 
  filter(!name %in% c("alaska", "hawaii"))
my_polisci_data <- my_polisci_data |> 
  filter(!state%in% c("alaska", "hawaii"))

static_ver <- states_polygon |>
  left_join(my_polisci_data, by = c("region" = "state"))
inter_ver <- states_sf |>
  left_join(my_polisci_data, by = c("name" = "state"))
```


## Static plot: Public opinion on pro-life

```{r}
#| echo: false

ggplot(static_ver, aes(long, lat, group = group, fill = prolife)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  coord_quickmap() +
  scale_fill_viridis_c(option = "C") +
  theme_void() +
  labs(
    fill = "Pro-life support",
    title = "State-level Support for Pro-life",
    caption = "Data is obtained from poliscidata R package"
  )
```


## Interactive plot: Public opinion on pro-life 

```{r}
#| echo: false

pal_num <- colorNumeric("plasma", inter_ver$prolife)

labels <- sprintf(
  "<strong>%s</strong><br/>Pro-life support: %s",
  str_to_title(inter_ver$name),
  inter_ver$prolife
) |> lapply(htmltools::HTML)

leaflet(inter_ver) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal_num(prolife),
    fillOpacity = 0.75,
    color = "white",
    weight = 1,
    label = labels,
    highlightOptions = highlightOptions(weight = 3, color = "#444")
  ) |>
  addLegend(
    pal = pal_num,
    values = ~prolife,
    title = "Pro-life support",
    position = "bottomright"
  )
```


## Static plot: Gay Support in the US

```{r}
#| echo: false

ggplot(static_ver, aes(long, lat, group = group, fill = gay_support3)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  coord_quickmap() +
  theme_void() +
  labs(
    fill = "Support level",
    title = "State-level Support for Gay Rights",
    caption = "Data is obtained from poliscidata R package"
  )
```


## Interactive plot: Gay Support in the US

```{r}
#| echo: false

pal_cat <- colorFactor("Set1", inter_ver$gay_support3)

labels_cat <- sprintf(
  "<strong>%s</strong><br/>Support: %s",
  str_to_title(inter_ver$name),
  inter_ver$gay_support3
) |> lapply(htmltools::HTML)

leaflet(inter_ver) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal_cat(gay_support3),
    fillOpacity = 0.8,
    color = "white",
    weight = 1,
    label = labels_cat,
    highlightOptions = highlightOptions(weight = 3)
  ) |>
  addLegend(
    pal = pal_cat,
    values = ~gay_support3,
    title = "Gay rights support",
    position = "bottomright"
  )
```

<br>

I definitely prefer the interactive plot. Even with making this conclusive closing, I felt that the interactive plot is very helpful with navigating the different states and also helped me to quickly identify which state is which. In places where gay supports are low (denoted by red), the prolife support is really high (denoted by yellow/brighter color). For example, states like Utah scores 61 in prolife support index and has low gay rights support.   